---
title: "You Zhang - Assignment 4"
output: html_notebook
---
In this homework, I will compare and analye the transponder flight data from 2015/04/04 and 2015/12/05.
Before you preview this notebook install the following packages (select and Run):

```{r, eval=FALSE}
install.packages("ggplot2")
install.packages("forcats")
install.packages("directlabels")
install.packages("ggthemes")
install.packages("ggiraph")
install.package('plotly')
install.package('dplyr')
```


The following codes will extract flights data from transponder flight logs to csv files. All data were downloaded from USC INF554 class resource.

```{r}
#create RTL15 data directory
dir.create('RTL15')
#will use dplyr and lubridate
library(dplyr)
library(lubridate)
library(ggplot2)
library(ggmap)
library(plotly)
library(dplyr)
library(sp)
library(geosphere)

######################################
# Loading data.
######################################

SMO <- data.frame(label = 'SMO', lon=-118.456667, lat=34.010167)
smo <- c(SMO$lon, SMO$lat)

# load RTL gzip file
# df <- load_rtl('./RTL15/RTL15040400.log.gz')

load_rtl <- function(filename) { 
  df <- read.csv(gzfile(filename), header=F)

  #V1 date: from 2015/04/03 to 2015/04/04
  #V2 is a time formatted as HH:MM:SS.000; from 16:00:00.000 to 15:59:58.000
  #V3 is 1111111
  #V4 is the transponder code (normally unique ICAO 24-bit address or (informally) Mode-S 'hex code')
  #V5 is the flight number
  #V6 is 'Unknown'
  #V7 is '0'
  #V8 == V9 altitude
  #V10 latitude
  #V11 longitude
  #V12-17 do not seem to be useful
  
  ######################################
  # Tidying-up the data.
  ######################################
  
  df <- df[,1:11]  #only keep first 11
  
  #set timestamp from V1(date) and V2(time)
  df <- df %>% mutate(timestamp = ymd_hms(paste(df$V1, df$V2), tz='America/Los_Angeles'))

  #remove not useful
  df$V1 <- NULL
  df$V2 <- NULL
  df$V3 <- NULL
  df$V6 <- NULL
  df$V7 <- NULL
  df$V9 <- NULL
  
  #rename columns
  names(df) <- c('code', 'flight', 'alt', 'lat', 'lon', 'timestamp')
  
  #drop records with invalid lat, lon and alt readings
  df <- df[df$lat != 0,]
  df <- df[df$alt > 0,]
  
  #fix flight names: strip white spaces from flight
  df$flight <- gsub('^[[:space:]]+|[[:space:]]+$', '', df$flight)
  df$flight[df$flight == '?'] <- ''
  df$flight[df$flight == '00000000'] <- ''
  df$flight[df$flight == '????????'] <- ''
  df$flight[df$flight == '@@@@@@@@'] <- ''
  
  #fix empty flights names with unique's flight associated using the code
  uniques <- unique(df[c('flight', 'code')])
  uniques <- uniques[uniques$flight != '',]
  na.flight <- which(df$flight == '')
  na.code <- df$code[na.flight]
  df$flight[df$flight == ''] <- uniques$flight[match(na.code, uniques$code, NA)]
  
  #order the dataframe
  df <- df[order(df$code, df$flight, df$timestamp, decreasing = FALSE),]
  
  #remove code starting with a60 and no flight name as these are from private planes.
  df <- df %>% filter(!(grepl('[acd]60', code) & is.na(flight)))
  df <- df %>% filter(!(grepl('[c]50', code) & is.na(flight)))
  df <- df %>% filter(!(grepl('[d]20', code) & is.na(flight)))
  
  #remove duplicate rows
  df <- unique(df)  #or with dplyr: df <- df %>% distinct()
  
  #using dplyr keep only first time for same flight position (alt, lon, lat)
  df <- df %>%
    group_by(code, flight, lat, lon, alt) %>% 
    summarize(timestamp = first(timestamp))
  
  #using dplyr keep only first alt for same flight location (lon, lat) and time
  df <- df %>%
    group_by(code, flight, lat, lon, timestamp) %>% 
    summarize(alt = first(alt))  #could use mean(alt)
  
  #using dplyr keep first alt, lat, lon for the same flight at the same time
  df <- df %>%
    group_by(timestamp, flight, code) %>%
    summarise(alt = first(alt), lat = first(lat), lon = first(lon))
  
  #add track variable to identify separate flights in the same day (delta(timestamp) > )
  df <- df[order(df$code, df$flight, df$timestamp, decreasing = FALSE),]  #ensure is ordered by flight & timestamp
  df <- df %>% group_by(code, flight) %>%
    mutate(track = as.integer(difftime(timestamp, lag(timestamp)))) %>%
    mutate(track = ifelse(is.na(track), 0, track)) %>%
    mutate(track = ifelse(track > 3600, 1, 0)) %>%  #3600s = 1h
    mutate(track = cumsum(track)) %>%
    mutate(id = paste0(code, flight, track))
  
  #count number of records per track
  df2 <- df %>% count(code, flight, track, id) %>% filter(n < 6)
  #df3 <- df %>% group_by(code, flight, track) %>% summarize(n = n())  #similar way of counting!
  
  #remove too few records per track
  df <- df %>% filter(!(id %in% df2$id))
  
  df <- df %>% arrange(flight, code, track, desc(timestamp))  #arrange rows with dplyr
  df <- df %>% select(timestamp, flight, code, track, id, lon, lat, alt)  #re-order column names
  
  #compute closest distance to SMOVOR (in meters)
  df$dist <- distHaversine(smo, cbind(df$lon, df$lat))  #geodesic distance (great-circle distance) to SMO

  # filter flights that get within 2Km from SMOVOR
  df_min <- df %>% filter(dist < 2000.0)  #closest observations to SMO
  df <- df %>% filter(id %in% df_min$id)  #observations that go within 2km of SMO.
  
  # keep only if alt between [3000; 11000] feet
  df <- df %>% filter(alt < 11000 & alt > 3000)  #observations that go within 2km of SMO.
  
  return (df)
}

# day 1: 150404
###############
df11 <- load_rtl('./RTL15/RTL15040400.log.gz');
df12 <- load_rtl('./RTL15/RTL15040500.log.gz');
df1 <- dplyr::bind_rows(df11, df12)

#filter night: (12AM-6:30AM) day: (6:30AM-12AM)
nighttime <- interval(ymd_hms('2015-04-04 00:00:00', tz='America/Los_Angeles'), ymd_hms('2015-04-04 06:30:00', tz='America/Los_Angeles'))
daytime <- interval(ymd_hms('2015-04-04 06:30:00', tz='America/Los_Angeles'), ymd_hms('2015-04-05 00:00:00', tz='America/Los_Angeles'))

df1d <- df1 %>% filter(timestamp %within% daytime)
df1n <- df1 %>% filter(timestamp %within% nighttime)
#verify time intervals with summary(df1d) and summary(df1n)

# save to csv for analysis
write.table(df1d, 'RTL150404_day.csv', sep = ',')
write.table(df1n, 'RTL150404_night.csv', sep = ',')

# day 2: 151205
###############
df11 <- load_rtl('RTL15/RTL15120500.log.gz');
df12 <- load_rtl('RTL15/RTL15120600.log.gz');
df1 <- dplyr::bind_rows(df11, df12)

#filter
#night: (12AM-6:30AM)
#day: (6:30AM-12AM)
nighttime <- interval(ymd_hms('2015-12-05 00:00:00', tz='America/Los_Angeles'), ymd_hms('2015-12-05 06:30:00', tz='America/Los_Angeles'))
daytime <- interval(ymd_hms('2015-12-05 06:30:00', tz='America/Los_Angeles'), ymd_hms('2015-12-06 00:00:00', tz='America/Los_Angeles'))

df1d <- df1 %>% filter(timestamp %within% daytime)
df1n <- df1 %>% filter(timestamp %within% nighttime)
#verify time intervals with summary(df1d) and summary(df1n)

# save to csv for analysis
write.table(df1d, 'RTL151205_day.csv', sep = ',')
write.table(df1n, 'RTL151205_night.csv', sep = ',')
```


There are four csv files created:
RTL150404_day, RTL150404_night, RTL151205_day, RTL151205_night
Each file contains 9 variables: timestamp, flight, code, track, id, longtitude, latitude, altitude and distance.

First, lets compare the different flight paths during in each time frame.

```{r}

df0404d <- df <- read.csv("RTL150404_day.csv", header=T)
df0404n <- df <- read.csv("RTL150404_night.csv", header=T)

df1205d <- df <- read.csv("RTL151205_day.csv", header=T)
df1205n <- df <- read.csv("RTL151205_night.csv", header=T)

SMO <- data.frame(label = 'SMO', lon=-118.456667, lat=34.010167)
smo <- c(SMO$lon, SMO$lat)

map.google <- get_map(smo, zoom = 10)  # get map around SMO
ggmap(map.google) +
  geom_point(data = SMO, aes(x = lon, y = lat), color="yellow", size=10, alpha=1) +
  geom_point(data = df0404d, 
             size=0.6, alpha=.3,
             aes(x = lon, y = lat, color=id, text = paste("Airline:", flight, "<br>code:", code, "<br>timestamp:", timestamp, "<br>alt:", alt))) +
  theme(legend.position = "none") +
  ggtitle("Scatterplot of Daytime Flights on 2015.04.04")

```
```{r}
SMO <- data.frame(label = 'SMO', lon=-118.456667, lat=34.010167)
smo <- c(SMO$lon, SMO$lat)

map.google <- get_map(smo, zoom = 10)  # get map around SMO
ggmap(map.google) +
  geom_point(data = SMO, aes(x = lon, y = lat), color="yellow", size=10, alpha=1) +
  geom_point(data = df1205d, 
             size=0.6, alpha=.3,
             aes(x = lon, y = lat, color=id, text = paste("Airline:", flight, "<br>code:", code, "<br>timestamp:", timestamp, "<br>alt:", alt))) +
  theme(legend.position = "none") +
  ggtitle("Scatterplot of Daytime Flight Paths on 2015.12.05")
```

As we can see in these two graphs, the number of flights has been significantly increased from April to December. Also, in April, most of the flights go through Los Angeles, while in December, there were much more flights going other directions.

In terms of night flights:
```{r}
SMO <- data.frame(label = 'SMO', lon=-118.456667, lat=34.010167)
smo <- c(SMO$lon, SMO$lat)

map.google <- get_map(smo, zoom = 10)  # get map around SMO
ggmap(map.google) +
  geom_point(data = SMO, aes(x = lon, y = lat), color="yellow", size=10, alpha=1) +
  geom_point(data = df0404n, 
             size=0.6, alpha=.3,
             aes(x = lon, y = lat, color=id, text = paste("Airline:", flight, "<br>code:", code, "<br>timestamp:", timestamp, "<br>alt:", alt))) +
  theme(legend.position = "none") +
  ggtitle("Scatterplot of Nighttime Flight Paths on 2015.04.04")
```
```{r}
SMO <- data.frame(label = 'SMO', lon=-118.456667, lat=34.010167)
smo <- c(SMO$lon, SMO$lat)

map.google <- get_map(smo, zoom = 10)  # get map around SMO
ggmap(map.google) +
  geom_point(data = SMO, aes(x = lon, y = lat), color="yellow", size=10, alpha=1) +
  geom_point(data = df1205n, 
             size=0.6, alpha=.3,
             aes(x = lon, y = lat, color=id, text = paste("Airline:", flight, "<br>code:", code, "<br>timestamp:", timestamp, "<br>alt:", alt))) +
  theme(legend.position = "none") +
  ggtitle("Scatterplot of Nighttime Flight Paths on 2015.12.05")

```

The number of flights also increased significantly. There were flights towards/from the west in December, comparing to April.

Now lets compare the altitude for both day.
First lets see the difference of night time flights. As it shows below, the average altitude in April normally higher than in December, but the number of flights in December is more than in April.

```{r}
df0404n$date<-"2015-04-04"
df1205n$date<-"2015-12-05"

dfnight <- bind_rows(df0404n,df1205n)

ggplot(dfnight) + 
  geom_bar(stat = "summary", fun.y = "mean", position=position_dodge(),
           aes(x=flight, y=alt,fill=date)) +
  ggtitle("Average Altitude Comparison (Nighttime)")
```

For the day time flights, flights in December has higher altitude than in April, and number of flights significantly increased in December as well.

```{r}
df0404d$date<-"2015-04-04"
df1205d$date<-"2015-12-05"

dfday <- bind_rows(df0404d,df1205d)

ggplot(dfday) + 
  geom_bar(stat = "summary", fun.y = "mean", position=position_dodge(),
             aes(x=flight, y=dist,fill=date)) +
  ggtitle("Average Distance Comparison (Daytime)")
```

Finally, lets see the difference of number of flights during different hours between the two days.

First for the daytime. Overall the number of flights increased in December. Especially in this graph, we can see that in April there were very few flights before 10am, while in December, morning flights has the largest amount during the day. In April, most of the flights are around 3pm, but in December there are few flights at the same time.

```{r}
library(lubridate)

dfday$hour <- hour(dfday$timestamp)

ggplot(dfday, aes(x = hour)) + 
  geom_histogram(aes(fill=date)) +
  labs(x="hour",y="number of flights") +
  ggtitle("Histogram of Hourly Flights Comparison (Daytime)")
```

Looking at the night time flights, there were no flights between 2am to 5am, while in December there were flights in that time frame.

```{r}
library(lubridate)

dfnight$hour <- hour(dfnight$timestamp)

ggplot(dfnight, aes(x = hour)) + 
  geom_histogram(aes(fill=date)) +
  labs(x="hour",y="number of flights") +
    ggtitle("Histogram of Hourly Flights Comparison (Nighttime)")
```

In conclusion, flight number increased a lot during year 2015. However, the average altitude decreased in the night time, and that's probably one of the reason why they increased the number of flights -- by flying lower, it's probably safer or shorter distances. Further researchs may require to confirm the ideas.
